Semantic Linearity in a Lazy Optimising Compiler

### 2024-02-28 icfp.tex
typo in S6
fixed usage env in 3.3

A \emph{usage environment} $\Delta$ is the means to encode the idea that lazy
bindings do not consume the resources required by the bound expression when
defined, but rather when the bindings themselves are consumed.
Specifically, we annotate so-called $\Delta$-bound variables with a \emph{usage
environment} to denote that consuming these $\D$-variables equates to consuming the
resources in the annotated usage environment $\D$ (which is essentially a
multiset of linear resources).
%
$\Delta$-bound variables are introduced by several constructs as (recursive)
let binders, case binders, and case pattern variables.
%
For example, the let-binder $u$ in the following program is annotated with a
usage environment $\{x,y\}$, the resources that are used in its body. Thus, in
the expression $e$, consuming $u$ will equate to consuming the linear resources
$x$ and $y$:
% , as per the insights developed in
% Section~\ref{sec:semantic-linearity-examples},
\[
f = \lambda \xl.~\lambda \y[1].~\llet{u = (x,y)}{e}
\]
Furthermore, usage environments guarantee that using a $\Delta$-bound variable
is mutually exclusive with directly using the resources it is annotated with --
using the $\Delta$-bound variable consumes all linear resources listed in its
usage environment, meaning they are no longer available for direct usage.
Dually, using the linear resources directly means they are no longer available
to consume through the usage environment of the $\Delta$-bound variable.

@@

\paragraph{Linear Haskell\label{sec:related-work-linear-haskell}.}
%
% \todo[inline]{Say something about Linear Haskell's lazy operational semantics,
% but the type system not being concerned with linearity in the presence of
% laziness}
%
% Haskell, contrary to most programming languages with linear types, has existed
% for 31 years of its life \emph{without} linear types. As such, the introduction
% of linear types to Haskell comes with added challenges that do not exist in
% languages that were designed with linear types from the start:
% %
% \begin{itemize}
%     \item Backwards compatibility. The addition of linear types shouldn't break
%         all existing Haskell code.
%     \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
%         its non-linearly-typed counterpart should fit in together and it must be
%         possible to define functions readily usable by both sides
%         simultaneously.
%     \item Future-proofing. Haskell, despite being an
%         industrial-strength language, is also a petri-dish for experimentation
%         and innovation in the field of programming languages. Therefore, Linear
%         Haskell takes care to accommodate possible future features, in
%         particular, its design is forwards compatible with affine and dependent
%         types.
% \end{itemize}
% %
Linear Haskell~\cite{cite:linearhaskell} augments the Haskell surface language
with linear types, but it is not concerned with extending GHC's intermediate
language(s), which presents its own challenges. 
%
However, the implementation Linear Haskell in GHC does modify and extend Core
with linearity/multiplicity annotations. Core's type system is unable to type
\emph{semantic} linearity of programs in the sense elaborated in our work,
which results in Core rejecting most linear programs resulting from optimising
transformations that leverage the non-strict semantics of Core.
%
Our work overcomes the limitations of Core's linear type system derived from
Linear Haskell by understanding linearity semantically in the presence of
laziness, and by proving linearity-preserving multiple Core-to-Core
optimisations GHC.
Linear Core can also be seen as a system that validates the programs written in
Linear Haskell and are compiled by GHC, by guaranteeing (through typing) that
linear resources are still used exactly once throughout the optimising
transformations.

>>>>

A \emph{usage environment} $\Delta$ is the means to encode the idea that lazy
bindings do not consume the resources required by the bound expression when
defined, but rather when the bindings themselves are consumed.
Specifically, we annotate so-called $\Delta$-bound variables with a \emph{usage
environment} to denote that consuming these $\D$-variables equates to consuming the
resources in the annotated usage environment $\D$ (which is essentially a
multiset of linear resources).
%
$\Delta$-bound variables are introduced by several constructs as (recursive)
let binders, case binders, and case pattern variables.
%
For example, the let-binder $u$ in the following program is annotated with a
usage environment $\{x{:}_1\s,y{:}_1\s\}$, the resources that are used in its body. Thus, in
the expression $e$, consuming $u$ will equate to consuming the linear resources
$x$ and $y$:
% , as per the insights developed in
% Section~\ref{sec:semantic-linearity-examples},
\[
f = \lambda \xl.~\lambda \y[1].~\llet{u_{\{x:_1\s,y:_1\s\}} = (x,y)}{e}
\]
Furthermore, usage environments guarantee that using a $\Delta$-bound variable
is mutually exclusive with directly using the resources it is annotated with --
using the $\Delta$-bound variable consumes all linear resources listed in its
usage environment, meaning they are no longer available for direct usage.
Dually, using the linear resources directly means they are no longer available
to consume through the usage environment of the $\Delta$-bound variable.

@@

\paragraph{Linear Haskell\label{sec:related-work-linear-haskell}.}
%
% \todo[inline]{Say something about Linear Haskell's lazy operational semantics,
% but the type system not being concerned with linearity in the presence of
% laziness}
%
% Haskell, contrary to most programming languages with linear types, has existed
% for 31 years of its life \emph{without} linear types. As such, the introduction
% of linear types to Haskell comes with added challenges that do not exist in
% languages that were designed with linear types from the start:
% %
% \begin{itemize}
%     \item Backwards compatibility. The addition of linear types shouldn't break
%         all existing Haskell code.
%     \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
%         its non-linearly-typed counterpart should fit in together and it must be
%         possible to define functions readily usable by both sides
%         simultaneously.
%     \item Future-proofing. Haskell, despite being an
%         industrial-strength language, is also a petri-dish for experimentation
%         and innovation in the field of programming languages. Therefore, Linear
%         Haskell takes care to accommodate possible future features, in
%         particular, its design is forwards compatible with affine and dependent
%         types.
% \end{itemize}
% %
Linear Haskell~\cite{cite:linearhaskell} augments the Haskell surface language
with linear types, but it is not concerned with extending GHC's intermediate
language(s), which presents its own challenges. 
%
However, the implementation of Linear Haskell in GHC does modify and extend Core
with linearity/multiplicity annotations. Core's type system is unable to type
\emph{semantic} linearity of programs in the sense elaborated in our work,
which results in Core rejecting most linear programs resulting from optimising
transformations that leverage the non-strict semantics of Core.
%
Our work overcomes the limitations of Core's linear type system derived from
Linear Haskell by understanding linearity semantically in the presence of
laziness, and by proving linearity-preserving multiple Core-to-Core
optimisations GHC.
Linear Core can also be seen as a system that validates the programs written in
Linear Haskell and are compiled by GHC, by guaranteeing (through typing) that
linear resources are still used exactly once throughout the optimising
transformations.
